Separating Use and Reuse to Improve Both

### 2017-05-25 FACS/Intro.tex
Revised RW

What the this-leaking problem shows is that adopting a more flexible
nominally typed OO model where subclassing does not imply subtyping is
not trivial, and a more substantial change in the language design is
necessary.  We are aware of at least 3 independently designed research
languages that address this limitation:
\begin{itemize}
\item In {\bf Delta-Trait (DT)}~\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282}
each  construct has a single responsability: classes instantiate objects,
interfaces induce types, records express state and traits are reuse units.
\item {\bf Package Templates (PT)}~\cite{KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12}:
are an extension of (full) Java where new packages can be ``syntetized'' by mixing
and integrating code templates. As an extension of java, PT allows but not requires
separation of inheritance and subtyping.
\item {\bf
    DeepFjig}~\cite{deep,servetto2014meta,DBLP:journals/iandc/LagorioSZ12} is
a module composition language where the main idea is that
nested classes with the same name are recursivelly composed.
\end{itemize}
\bruno{We need here a better motivation. From the reviewer perspactive
he will think: ``So, there are already 3 solutions, why should I care
about a new one?''}
This paper aims at showing a simple language design, called \name, to
completely decouple subtyping and subclassing in a nominally typed OO
language. The key idea is to divide between code designed for
\textbf{USE} and code designed for \textbf{REUSE}. 
In \name there are two separate concepts: classes
and traits~\cite{Traits:ECOOP2003}. Classes are meant for code use, and cannot be used
for reuse. In some sense classes in \name are like final classes in
Java. Traits are meant for code reuse and multiple traits can be
composed to form a class which can then be instantiated. Traits 
cannot be instantiated (or used) directly. Such design allows the
subtyping and code reuse to be treated separately, which in turn
brings several benefits in terms of flexibility and code reuse.
In summary, our contributions are:

>>>>

What the this-leaking problem shows is that adopting a more flexible
nominally typed OO model where subclassing does not imply subtyping is
not trivial, and a more substantial change in the language design is
necessary.  We are aware of at least 3 independently designed research
languages that address this limitation:
\begin{itemize}
\item In {\bf Delta-Trait (DT)}~\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282}
each  construct has a single responsability: classes instantiate objects,
interfaces induce types, records express state and traits are reuse units.
\item {\bf Package Templates (PT)}~\cite{KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12}:
are an extension of (full) Java where new packages can be ``syntetized'' by mixing
and integrating code templates. As an extension of java, PT allows but not requires
separation of inheritance and subtyping.
\item {\bf
    DeepFjig}~\cite{deep,servetto2014meta,fjig} is
a module composition language where the main idea is that
nested classes with the same name are recursivelly composed.
\end{itemize}
\bruno{We need here a better motivation. From the reviewer perspactive
he will think: ``So, there are already 3 solutions, why should I care
about a new one?''}
This paper aims at showing a simple language design, called \name, to
completely decouple subtyping and subclassing in a nominally typed OO
language. The key idea is to divide between code designed for
\textbf{USE} and code designed for \textbf{REUSE}. 
In \name there are two separate concepts: classes
and traits~\cite{Traits:ECOOP2003}. Classes are meant for code use, and cannot be used
for reuse. In some sense classes in \name are like final classes in
Java. Traits are meant for code reuse and multiple traits can be
composed to form a class which can then be instantiated. Traits 
cannot be instantiated (or used) directly. Such design allows the
subtyping and code reuse to be treated separately, which in turn
brings several benefits in terms of flexibility and code reuse.
In summary, our contributions are:

### 2017-05-25 FACS/RelatedWork.tex
Revised RW

\saveSpace\saveSpace
\section{Related Work}
\saveSpace\saveSpace
Literature on code reuse is too vast to let us do justice of it in fiew pages,
Our work is inspired on Traits~\cite{ducasse2006traits}; they are in turn insipired on module composition languages~\cite{ancona2002calculus}.
\saveSpace
\subsection{Inheritance and subtyping}
\saveSpace
We are aware of at least 3 independently designed research languages 
that address this limitation:Delta-Trait(DT)\cite{ferruccioworks}, Package Template(PT)\cite{}, DeepFjig\cite{}.
Leveraging on \emph{traits}, in this work we aim to synthesizes
the main ideas of those very different designs.

\begin{itemize}
\item DeepFjig+ {\bf A simple uniform syntax for code literals}
DeepFjig is best in this sense, since DT have separate syntax for class literals, trait literals and record literals.
PT on the other side build on top of full Java, thus has a very involved syntax.
Thanks to our novel represantation of state, our approach offers a much simpler and uniform syntax then
all others, where evertyhing is basically just a method.
\item DeepFjig-
{\bf Reusable code can not be ``used'', that is instantiated or used as a type.}
This happens in DT and in PT, but not in DeepFjig. To allow reusable code to be directly 
usable, in DeepFjig
classes introduce nominal types in an innatural way: the type of \Q@this@ is only \Q@This@ (sometimes called \Q@<>@) and not the nominal type of its class: that is in DeepFjig 
\Q@A:{ method A m()this}@ is not well typed. This is because
\Q@B: Use A@ flattens to \Q@B:{ method A m()this}@, that is clearly not well typed.
Looking to this example is clear why we need reusable code to be agnostic of its name.
Then, either reusable code have no name (DT, PT and this work)
or all code is reusable and usable, and all code need to be auckwardly agnostic of its name.

\item PT+
{\bf Requiring abstract signatures is a left over of module composition mindset.}
DT and Deepfjig comes form a tradition of functional module composition, where 
modules are typed in isolation under an environments, and then the composition is performed.
As we show in this work, this is ends up requiring verbose repetition of abstract signatures, 
that for hightly modularized code may end up composing most of the program.
Simple Java (and thus PT, since it is an add on over Java) show us a better way:
the meaning of names can be understood from the reuse context.
The typing strategy of PT offers the same advantages of our typing model, 
but is more involved and indirect. This may be caused by the
heavy task of integrating with full Java.
\item PT-{\bf Composition algebra.}
The idea of using composition operators over atomic values as in an arithmetic expression is very powerfull,
and makes easy to extend languages with more operators. Deepfjig and DT embrace this idea, while PT takes the traditional Java/C++ approach of using enhanced class/package declaration syntax.
The typing strategy of PT also seams to be connected with this decision, so it would be hard to move their approach in a composition algebra setting.
\item DT- {\bf Naming your type, even if you have none yet.}
Both Deepfjig and PT allows a class to refer to its name, abait is more unobvious in PT since you have to introduce both a package and a classes to express it.
This allows to encode binary methods, express patterns like withers or fluent setters and to instantiate instances of the (future) class(es)  using the reused code.
\item DT+{\bf Complete ontological separation between use and reuse}
While all 3 works allows to separate inheritance and subtyping only DT properly enforce 
separation between use (classes and interfaces) and reuse (traits).
This is because in DeepFjig all classes are both units of use and reuse (however, subtype is not induced)
and PT imports all the complexity of Java, so abait is possible to separate use and reuse, the model have powerfull but unobvous implications where (conventional Java) \Q@extends@ and PT are used together

\end{itemize}
With the exception of those 3 lines of work, to the best of our understading
other famous work in literature, like~\cite{odersky2008programming,nystrom2006j}
does no completly break the relation between inheritance and subtyping, but only prevent subtyping only where 
it would be unsound.

@@

The idea of abstract state operation has emerged from CJ~\cite{wang2016classless}, and offer a clean solution to handling state
in a trait composition setting.
Note how abstract state operation are different from just hiding fields under getter and setters: 
in our model the programmer simply never declare what is the state of the class, not even what information is stored in fields.
The state is computed by the system as an overall result of the whole code composition process.

In literature there has been many attempts to add state in traits/module composition languages.
\begin{itemize}  
\item No constructor: all the fields start at null/a default specified value.
  Fields are just like another kind of (abstract) member, and two fields
  with identical type can be merged by sum/use; \Q@new C()@ can be used for all classes, and \Q@init@ methods may be called later, as in
  \Q@Point p=new Point(); p.init(10,30)@.

  To its credit, this simple approach is commutative and associative and do not disrupt elegance of summing methods.
  However, objects are created "broken" and the user is trusted with fixing them.
  While is easy to add fields, the load of inizializing them is on the user; moreover
    all the objects are intrinsically mutable, so this model is unfriendly
    to a functional programming style.
\item Constructor compose fields:
In this approach (used by \cite{FJig}) the fields are declared but not initialized, and
a canonical constrctor constructor (as in FJ) taking a value for each field and just initializing such field
is automatically generated in the resulting class.

It is easy to add fields, however this model is associative but not commutative: composition order influence field order, and thus the construtor signature.
Self construction is not possible 
since the signature of the constuctor change during composition.

\item Constructor can be composed if they offers the same exact parameters:
In this approach (used by DeepFJig) traits declare field and constructors.
The constructor initialize the fields but can do any other computation.
Traits whose constructors have the same signature can be composed.
The composed constructor will execute both constructor bodies in order.

This approach is designed to allows Self construction.
It is also associative and mostly commutative: composition order only influence execution order of side effects during construction.
However constructors composition requires identical constructor signature: this
hamper reuse, and if a field is added, its initial value need to be
somehow syntetized from the constructor parameters.

>>>>

\saveSpace\saveSpace
\section{Related Work}
\saveSpace\saveSpace
Literature on code reuse is too vast to let us do justice of it in a few pages.
Our work is inspired by traits~\cite{ducasse2006traits}, which in turn
are inspired by module composition languages~\cite{ancona2002calculus}.
\saveSpace
\subsection{Inheritance and subtyping}
\saveSpace
We are aware of at least 3 independently designed research languages 
that address the this-leaking problem: Delta-Trait~(DT)\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282}, Package Templates~(PT)\cite{KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12}, DeepFjig\cite{deep,servetto2014meta,fjig}.
Leveraging on \emph{traits}, in this work we aim to synthesize
the best ideas of those very different designs, while at the same time 
coming up with a simpler and improved design for separating
subclassing from subtyping. The following compares 
various aspects of the language designs:
\bruno{can we say that by borrowing ideas from various designs we come
up with a simpler design, which also addresses various limitations of
particular designs?}

\begin{itemize}
\item DeepFjig+ {\bf A simple uniform syntax for code literals}
DeepFjig is best in this sense, since DT has separate syntax for class literals, trait literals and record literals.
PT on the other hand is built on top of full Java, thus has a very
involved syntax.
\name leverages on DeepFjig's approach but,
\emph{thanks to our novel representation of state}, \name also offers a much simpler and uniform syntax than
all other approaches: everyhing is just a method.
\item DeepFjig-
{\bf Reusable code cannot be ``used'', that is instantiated or used as a type.}
This happens in DT and in PT, but not in DeepFjig. To allow reusable code to be directly 
usable, in DeepFjig
classes introduce nominal types in an unnatural way: the type of
\Q@this@ is only \Q@This@ (sometimes called \Q@<>@) and not the
nominal type of its class. 
That is in DeepFjig 
\Q@A:{ method A m()this}@ is not well typed. This is because
\Q@B: Use A@ flattens to \Q@B:{ method A m()this}@, which is clearly not well typed.
Looking to this example is clear why we need reusable code to be agnostic of its name.
Then, either reusable code has no name (as in DT, PT and \name)
or all code is reusable and usable, and all code needs to be awkwardly agnostic of its name.

\item PT+
{\bf Requiring abstract signatures is a left over of module composition mindset.}
DT and Deepfjig comes from a tradition of functional module composition, where 
modules are typed in isolation under an environment, and then the composition is performed.
As we show in this work, this ends up requiring verbose repetition of abstract signatures,
which (for highly modularized code) may end up constituting most of the program.
Simple Java (and thus PT, since it is a Java extension) shows us a better way:
the meaning of names can be understood from the reuse context.
The typing strategy of PT offers the same advantages of our typing model, 
but is more involved and indirect. This may be caused by the
heavy task of integrating with full Java.
\item PT-{\bf Composition algebra.}
The idea of using composition operators over atomic values as in an arithmetic expression is very powerful,
and makes it easy to extend languages with more operators. Deepfjig and DT embrace this idea, while PT takes the traditional Java/C++ approach of using enhanced class/package declaration syntax.
The typing strategy of PT also seems to be connected with this decision, so it would be hard to move their approach in a composition algebra setting.
\item DT- {\bf Naming your type, even if there is none yet.}
Both Deepfjig and PT allow a class to refer to its name, albeit this is
less obvious in PT since both a package and a class have to be introduced to express it.
This allows encoding binary methods, express patterns like withers or fluent setters and to instantiate instances of the (future) class(es)  using the reused code.
\item DT+{\bf Complete ontological separation between use and reuse}
While all 3 works allow separating inheritance and subtyping only DT properly enforces 
separation between use (classes and interfaces) and reuse (traits).
This is because in DeepFjig all classes are both units of use and reuse (however, subtyping is not induced).
PT imports all the complexity of Java, so although is possible to separate use and reuse, the model have powerful but non-obvious implications where (conventional Java) \Q@extends@ and PT are used together.

\end{itemize}
With the exception of those 3 lines of work, to the best of our understading
other famous work in literature, like~\cite{odersky2008programming,nystrom2006j}
does not completely break the relation between inheritance and subtyping, but only prevents subtyping where 
it would be unsound.\bruno{structural typing? Cook?}

@@

The idea of abstract state operations emerged from Classless
Java~\cite{wang2016classless}. This approach offers a clean solution to handle state
in a trait composition setting.
Note how abstract state operations are different from just hiding fields under getter and setters: 
in our model the programmer simply never has to declare what is the state of the class, not even what information is stored in fields.
The state is computed by the system as an overall result of the whole code composition process.

In the literature there have been many attempts to add state in traits/module composition languages:
\begin{itemize}  
\item {\bf No constructor}: all the fields start with {\bf null} or a default specified value.
  Fields are just like another kind of (abstract) member, and two fields
  with identical types can be merged by sum/use; \Q@new C()@ can be used for all classes, and \Q@init@ methods may be called later, as in
  \Q@Point p=new Point(); p.init(10,30)@.

  To its credit, this simple approach is commutative and associative and does not disrupt elegance of summing methods.
  However, objects are created "broken" and the user is trusted with fixing them.
  While is easy to add fields, the load of inizializing them is on the user; moreover
    all the objects are intrinsically mutable, so this model is unfriendly
    to a functional programming style.
\item {\bf Constructors compose fields}:
In this approach (used by \cite{fjig}) the fields are declared but not initialized, and
a canonical constructor constructor (as in FJ) taking a value for each field and just initializing such field
is automatically generated in the resulting class.
It is easy to add fields, however this model is associative but not commutative: composition order influences field order, and thus the construtor signature.
Self construction is not possible 
since the signature of the constuctors changes during composition.

\item {\bf Constructors can be composed if they offer the same exact parameters}:
In this approach (used by DeepFJig) traits declare fields and constructors.
The constructor initializes the fields but can do any other computation.
Traits whose constructors have the same signature can be composed.
The composed constructor will execute both constructor bodies in order.
This approach is designed to allows Self construction.
It is also associative and mostly commutative: composition order only influences execution order of side effects during construction.
However constructor composition requires identical constructor signatures: this
hampers reuse, and if a field is added, its initial value needs to be
somehow syntetized from the constructor parameters.
